Evolution of ORM Systems: Formal Foundations

### 2013-01-17 doc/description.tex
updated definition of mapping between instances

@@

\paragraph{Copy Value} The function copy values from one column in a table to another column in the same or in another table - according to mapping function $f_m$. 
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc}}{\begin{gathered}
			insertValues(second(p), \\ [l_{tc},value(findPair(l_{sc}, first(p)))],d)
		\end{gathered}}
 	\end{cases} \\
 	l_s, l_t, l_{sc}, l_{tc} \in Label, d \lhd Database
\end{align*}

\paragraph{Copy rows} The function $copyRows$ creates a copy of a source table content in another table with the same structure. 
\begin{align*}
	\mathbf{copyRows} : Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyRows(l_s, l_t, f_m, d) = \begin{cases}
 	\inference{|f_m| = \varnothing}{d} \\\\
 	\inference{|f_m| = 1 \\ \wedge \exists t_s, t_t \in tables(d) : name(t_s) = l_s, name(t_t) = l_t \\ \wedge t_s \cong t_t }{data(d) \cup insertRow(row(l, pairs(first(f_m))), d)} \\\\
 	\inference{|f_m| > 1}{\begin{gathered}
		 d \bullet copyRows(l_s, l_t, head(f_m)) \bullet copyRows(l_s, l_t, tail(f_m))
	\end{gathered}}
  \end{cases} \\
  l_s, l_t \in Label, d \lhd Database
\end{align*}

@@

\paragraph{Add association} Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be paired. This pairing is provided by the function $f_m$:
$$f_m : \rightarrow Row \times Row $$

\textbf{TODO: define the features of the $f_m$ function}
\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, f_m)) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, f_m)) = \begin{cases}
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l\\ 		\wedge name(t_t) = reference(a) \\ \wedge \forall f \in foreignKeys(t_t) \cuo columns(t_t) : name(f) \neq name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
		database(s) \bullet addForeignKey(name(t_s), \rho(a,d), d) \\ \bullet \forall r \in f_m(l, reference(a)) : \\ insertRow(row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]), d)
	\end{gathered}}
 \\ \\
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l \\ \wedge name(t_t) = reference(a) \\ \wedge startCardinality(a) > 1}{\begin{gathered} 
	database(s) \bullet addTable(\rho(a,d),d) \\ \bullet \forall r \in f_m(l, reference(a)) :\\ insertRow(row(name(owningClass(a)) + name(a), \\ [name(t_s), value(primaryKey(first(r)))]\\ [name(t_t), value(primaryKey(second(r)))]))
	\end{gathered}} 
\end{cases}\\
\end{align*}

@@

%\begin{align*}
%insertClass : Class \times Database \rightarrow Database \\
%insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
%	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
%	assocaiations(c) = \varnothing
%}{
%\begin{gathered}
% addTable(\rho(c, d),d) \\ \wedge t \rho(singleProperties(c)) \wedge \rho(collections(c)) \\ 
%\end{gathered}
%}
%\end{align*}

@@

The $CopyProperty$ function can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, therefore transformation to update data are used.
\begin{align*}
	\mathbf{CopyProperty} : Label \times Label \times Label \times (Row \times Row) \times State \rightarrow State \\
	\Psi(CopyProperty(l_s, l_t, l_p, f_m), s) = \inference{\exists c \in classes(s) : name(c) = l_s \\ \wedge \exists p \in properties(c) : name(p) = l_p }{application(s) \circ addProp(l_t, p)} 
	\\\\
	\Phi(CopyProperty(l_s, l_t, l_p, f_m), s) = \begin{cases}
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) \leq 1 }{\begin{gathered}
			database(s) \bullet copyColumn(l_s, l_t, l_p) \\ \bullet copyValues(l_s, l_t, l_p, l_p, f_m) 
			\end{gathered}} \\ \\
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) > 1}{\begin{gathered}
			database(s) \bullet copyTable(l_s + l_p, l_t + l_p) \\ \bullet copyRows(l_s + l_p, l_t + l_p, f_m) 
			\end{gathered}}
 \end{cases}
\end{align*}

With the copy property transformation defined we are able to define advanced evolution scenarios.

>>>>

\paragraph{Mapping of Instances} The set $M$ which contains all possible mappings between classes, has to be defined to allow several evolutionary operations. Each mapping represents an ordered set of pairs of instances. These pairs can represent one-to-one (or many) relationship or many-to-many relationship between instances.
\begin{align*}
	m_{l_1, l_2} =  <(r_1, r_2)*>,\\
	r_1 \in tableData(l_1, database(s)) \Rightarrow \exists p \in m_{l_1, l_2} \wedge first(p) = r_1 \\
	r_2 \in tableData(l_2, database(s)) \Rightarrow \exists p \in m_{l_1, l_2} \wedge second(p) = r_2\\
	l_1, l_2 \in Label, r_1, r_2 \lhd Row
\end{align*}
The definition does not allow to add any default value in case the number of instances does not match. It is because the evolution transformations are designed to not affect stored data (instances) by creating a new data which does not origin from the software.
%The features of both sets allows only pairs from existing instances and it assumes there is always at least one pair for each row from both tables.

@@

\paragraph{Copy Value} The function copy values from one column in a table to another column in the same or in another table - according to provided mapping. 
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} o_{l_s, l_t}, d) = \begin{cases}
		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc}}{\begin{gathered}
		\forall	p \in m_{l_s, l_t} :	d \bullet insertValues(second(p), \\ [l_{tc},value(findPair(l_{sc}, first(p)))])
		\end{gathered}}
 	\end{cases} \\
 	l_s, l_t, l_{sc}, l_{tc} \in Label, d \lhd Database
\end{align*}

\paragraph{Copy rows} The function $copyRows$ creates a copy of a source table content in another table with the same structure. 
\begin{align*}
	\mathbf{copyRows} : Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyRows(l_s, l_t, m_{l_s, l_t}, d) = \begin{cases}
 	\inference{|m_{l_s, l_t}| = \varnothing}{d} \\\\
 	\inference{|m_{l_s, l_t}| = 1 \\ \wedge \exists t_s, t_t \in tables(d) : name(t_s) = l_s, name(t_t) = l_t \\ \wedge t_s \cong t_t }{d \bullet insertRow(row(l, pairs(first(m_{l_s, l_t}))), d)} \\\\
 	\inference{|m_{l_s, l_t}| > 1}{\begin{gathered}
		 d \bullet copyRows(l_s, l_t, head(M_{l_s, l_t})) \\ \bullet copyRows(l_s, l_t, tail(m_{l_s, l_t}))
	\end{gathered}}
  \end{cases} \\
  l_s, l_t \in Label, d \lhd Database, m_{l_s, l_t} \in M
\end{align*}

@@

\paragraph{Add association} Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be connected on the database level via foreign key or connection table. The information about pairing has to be provided by the user in form of set of instances pairs, because the model does not contain enough information. 
\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, m_{l, reference(a)})) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, m_{l, reference(a)})) = \begin{cases}
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l\\ 		\wedge name(t_t) = reference(a) \\ \wedge \forall f \in foreignKeys(t_t) \cup columns(t_t) \\: name(f) \neq name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
		database(s) \bullet addForeignKey(name(t_s), \rho(a,d), d) \\ \bullet \forall r \in m_{l, reference(a)} : \\ insertRow(row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]), d)
	\end{gathered}}
 \\ \\
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l \\ \wedge name(t_t) = reference(a) \\ \wedge startCardinality(a) > 1}{\begin{gathered} 
	database(s) \bullet addTable(\rho(a,d),d) \\ \bullet \forall r \in m_{l, reference(a)} :\\ insertRow(row(name(owningClass(a)) + name(a), \\ [name(t_s), value(primaryKey(first(r)))]\\ [name(t_t), value(primaryKey(second(r)))]))
	\end{gathered}} 
\end{cases}\\
\end{align*}

@@

@@

The $CopyProperty$ function can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, therefore transformation to update data are used.
\begin{align*}
	\mathbf{CopyProperty} : Label \times Label \times Label \times (Row \times Row) \times State \rightarrow State \\
	\Psi(CopyProperty(l_s, l_t, l_p, m_{l_s, l_t}), s) = \inference{\exists c \in classes(s) : name(c) = l_s \\ \wedge \exists p \in properties(c) : name(p) = l_p }{application(s) \circ addProp(l_t, p)} 
	\\\\
	\Phi(CopyProperty(l_s, l_t, l_p, m_{l_s, l_t}), s) = \begin{cases}
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) \leq 1 }{\begin{gathered}
			database(s) \bullet copyColumn(l_s, l_t, l_p) \\ \bullet copyValues(l_s, l_t, l_p, l_p, m_{l_s, l_t}) 
			\end{gathered}} \\ \\
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) > 1}{\begin{gathered}
			database(s) \bullet copyTable(l_s + l_p, l_t + l_p) \\ \bullet copyRows(l_s + l_p, l_t + l_p, m_{l_s, l_t}) 
			\end{gathered}}
 \end{cases}\\
 l_s, l_t, l_p \in Label,  m_{l_s, l_t}) \in M, s \lhd Software
\end{align*}
With the copy property transformation defined we are able to define advanced evolution scenarios.
